Longitudinal Analysis Manuscript: Working Draft

### 2023-04-21 test.qmd
cli commit 10

---
title: "Longitudinal Analysis Manuscript: Working Draft"
subtitle: Longitudinal Analysis using data from the ABCD Study 
authors:
  - name: Samuel W. Hawes
    department: Center for Children and Families
    affiliation: Florida International University
    location: Miami, FL
    email: shawes@fiu.edu
  - name: Andrew K. Littlefield
    department: Psychological Sciences
    affiliation: Texas Tech University
    location: Lubbock, TX
    email: andrew.littlefield@ttu.edu
  - name: Daniel A. Lopez
    department: Department of Public Health
    affiliation: University of Rochester Medical Center
    location: Rochester, NY
    email: Daniel_Lopez2@URMC.Rochester.edu
  - name: Wesley K. Thompson
    department: Center for Population Neuroscience and Genetics
    affiliation: Laureate Institute for Brain Research
    location: Tulsa, OK
    email: wes.stat@gmail.com
  - name: Kenneth J. Sher
    department: Psychological Sciences
    affiliation: University of Missouri
    location: Columbia, MO
    email: SherK@missouri.edu
  - name: Additional Co-authors
    department: xxxxxx
    affiliation: University of xxxx
    location: xxxxxxx
    email: xxxxxx

@@

## Modeling Data Across Two Time Points versus Three or More Time Points.
Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions. Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1). At the same time, data based on two assessments are inherently limited on multiple fronts. As [@rogosa1982] noted approximately forty years ago, “Two waves of data are better than one, but maybe not much better”. These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for model identification and accurate parameter estimation. It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (since linear, straight line change is the only possible form, given two assessments; see [@duncan2009]). Research designs that include three or more time points allow for increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of inferential research. To illustrate, developmental theories are typically interested in understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level. This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield20210603. It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling. Indeed, several contemporary longitudinal models designed to isolate within-individual variability [e.g., the Latent Curve Model with Structured Residuals; @curran2014a] require at least four assessments to parameterize fully and, more generally, increasingly accurate parameter estimates are obtained as more assessment occasions are used [@duncan2009].

@@

There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021. Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011]. Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. As a result, experts increasingly caution about the need to xxxxxxxx [add citation]. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield20210603; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a]. It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield20210603, for further discussion].

## Model Assumptions
Many statistical models assume certain characteristics about the data to which they are being applied. As an example, common assumptions of parametric statistical models include normality, linearity, and equality of variances. These assumptions must be carefully considered before conducting analysis so that valid inferences can be made from the data; that is, violation of a model’s assumptions can substantively alter the interpretation of results. Similarly, statistical models employed in the analyses of longitudinal data often entail a range of assumptions that must be closely inspected. One central issue for repeated measurements on an individual is how to account for the correlated nature of the data; another common feature of longitudinal data is heterogeneous variability; that is, the variance of the response changes over the duration of the study. Traditional techniques, such as a standard regression or ANOVA model, assume residuals are independent and thus are inappropriate for designs that assess (for example) the same individuals across time. That is, given the residuals are no longer independent, the standard errors from the models are biased and can produce misleading inferential results. Although there are formal tests of independence for time series data (e.g., the Durbin-Watson statistic; Durbin & Watson, 1950), more commonly independence is assumed to be violated in study designs with repeated assessments. Therefore, an initial question to be addressed by a researcher analyzing prospective data is how to best model the covariance structure of said data.

@@

>>>>

---
title: "Longitudinal Analysis Manuscript: Working Draft"
subtitle: Longitudinal Analysis using data from the ABCD Study 
authors:
  - name: Samuel W. Hawes
    department: Center for Children and Families
    affiliation: Florida International University
    location: Miami, FL
    email: shawes@fiu.edu
  - name: Andrew K. Littlefield
    department: Psychological Sciences
    affiliation: Texas Tech University
    location: Lubbock, TX
    email: andrew.littlefield@ttu.edu
  - name: Daniel A. Lopez
    department: Department of Public Health
    affiliation: University of Rochester Medical Center
    location: Rochester, NY
    email: Daniel_Lopez2@URMC.Rochester.edu
  - name: Kenneth J. Sher
    department: Psychological Sciences
    affiliation: University of Missouri
    location: Columbia, MO
    email: SherK@missouri.edu
  - name: Additional Co-authors
    department: xxxxxx
    affiliation: University of xxxx
    location: xxxxxxx
    email: xxxxxx
  - name: Wesley K. Thompson
    department: Center for Population Neuroscience and Genetics
    affiliation: Laureate Institute for Brain Research
    location: Tulsa, OK
    email: wes.stat@gmail.com

@@

## Modeling Data Across Two Time Points versus Three or More Time Points.
Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions. Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1). At the same time, data based on two assessments are inherently limited on multiple fronts. As [@rogosa1982] noted approximately forty years ago, “Two waves of data are better than one, but maybe not much better”. These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for model identification and accurate parameter estimation. It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (since linear, straight line change is the only possible form, given two assessments; see [@duncan2009]). Research designs that include three or more time points allow for increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of inferential research. To illustrate, developmental theories are typically interested in understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level. This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield2021. It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling. Indeed, several contemporary longitudinal models designed to isolate within-individual variability [e.g., the Latent Curve Model with Structured Residuals; @curran2014a] require at least four assessments to parameterize fully and, more generally, increasingly accurate parameter estimates are obtained as more assessment occasions are used [@duncan2009].

@@

There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021. Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011]. Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. As a result, experts increasingly caution about the need to xxxxxxxx [add citation]. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a]. It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].

## Model Assumptions
Many statistical models assume certain characteristics about the data to which they are being applied. As an example, common assumptions of parametric statistical models include normality, linearity, and equality of variances. These assumptions must be carefully considered before conducting analysis so that valid inferences can be made from the data; that is, violation of a model’s assumptions can substantively alter the interpretation of results. Similarly, statistical models employed in the analyses of longitudinal data often entail a range of assumptions that must be closely inspected. One central issue for repeated measurements on an individual is how to account for the correlated nature of the data; another common feature of longitudinal data is heterogeneous variability; that is, the variance of the response changes over the duration of the study. Traditional techniques, such as a standard regression or ANOVA model, assume residuals are independent and thus are inappropriate for designs that assess (for example) the same individuals across time. That is, given the residuals are no longer independent, the standard errors from the models are biased and can produce misleading inferential results. Although there are formal tests of independence for time series data (e.g., the Durbin-Watson statistic [@durbin1950]), more commonly independence is assumed to be violated in study designs with repeated assessments. Therefore, an initial question to be addressed by a researcher analyzing prospective data is how to best model the covariance structure of said data.

@@

# Part III Section in Progress: TBD
\label{sec:headings}
**Subsection TBD.**

# Part IV Discussion
\label{sec:headings}
**Subsection TBD.**
